Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-07-17 ms.Rnw
progress

\section{From population processes to genealogy processes}

@@

@@

\paragraph{Pruning}

@@

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $j\in\Zp$ and $t\in\Rp$, $Y^{P}_j(t)=\left(\ct^{P}_j(t),\anc^{P}_j(t),\deme^{P}_j(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $Y:\Rp\to\Yspace$ defines a pruned genealogy.

@@

\paragraph{Lineage count, saturation}

@@

\subsection{Obscured genealogy}

@@

\begin{thm}\label{thm:pruned_lik}
  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
  Then
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y(e))}{r^{U_e}}{s(\yt(e),y(e))}\,Q_{U_e}(\yt(e),y(e))}.
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*}=0$.
  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $\U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, one ends up with $s_i$ black balls, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Since our genealogies are labelled, we are interested in the probability that the specific lineages are involved.
  We therefore divide by $\binom{\ell_i}{s_i}$.
  Since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
  \end{equation*}
  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

\begin{equation*}
  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
\end{equation*}

@@

\begin{thm}\label{thm:obsc_lik}
  Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$ for all $u\in\Jumps$, $t\in\Rp$, $\xi,\xi'\in\Xspace$, and $\eta\in\Yspace$.
  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ if and only if $Q_u(\eta,\eta')=1$.
  Let $\rdm{y}_t$ be the Markov jump process generated by the kernel $\pi$, with jump times fixed at $\event(\Hist_t)$.
  Then
  \begin{equation*}
    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_t,\rdm{y}_t)}}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
  \end{equation*}
  the expectation being taken over $\rdm{y}_t$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \Prob{V^*\vert\Hist_t}=\int{\Indicator{\obs(P)=V^*}\,\Prob{P\vert\Hist_t}\,\dd{P}},
  \end{equation}
  the integral being taken over all pruned genealogies.
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $V^*$ as sample paths from a Markov process on $\Yspace$ and
  treat the integral in \cref{eq:IS1} as an expectation over these paths.
  Conditional on $\Hist_t$, the probability kernel $\pi$ generates a Markov chain, $\rdm{\hat{y}}_k$, on $\Yspace$ such that
  \begin{equation*}
    \Prob{\rdm{\hat{y}}_k=\hat{y}_k\vert\rdm{\hat{y}}_{k-1}=\hat{y}_{k-1}}=\pi_{U_{t_k}}(t_k,\leftlim{X}_{t_k},X_{t_k},\hat{y}_{k-1},\hat{y}_k).
  \end{equation*}
  $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
  The conditions on $\pi$ guarantee that for every $P$ such that $\obs(P)=V^*$, then $Y^P$ is a sample path of this process and, conversely, that for every sample path $y$, there is a pruned genealogy $P=P(y)$ such that $\obs(P)=V^*$ and $Y^P(t)=y_t$.
  In particular, they guarantee that $\pi_u(\xi,\xi',\eta,\eta')\propto{Q_u(\eta,\eta')}$.

  We can therefore write
  \begin{equation*}
    \Prob{V^*\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}},
  \end{equation*}
  the expectation being taken with respect to the random process $\rdm{y}_t$.
  Here, by definition,
  \begin{equation*}
    \pi(\rdm{y}\vert\Hist_t)=\prod_{e\in\event(\Hist_t)}{\pi_{U_{e}}(e,\leftlim{X}_{e},X_{e},\leftlim{\rdm{y}}_{e},\rdm{y}_{e})}.
  \end{equation*}
  The result follows from \cref{thm:pruned_lik}.
\end{proof}

\AAK{Two approaches possible:}
\begin{enumerate}
\item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
  In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
\item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
  That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
  Therefore, any importance sampling scheme for $\Prune_t$ will do.
  In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
  We show that this is equivalent to the specific filter equation.

  It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
  Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
\end{enumerate}

@@

\paragraph{Integrating out the history, filter equations}

@@

\begin{defn}
  Suppose $X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
  \begin{equation}\label{eq:kfe2}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}},
  \end{equation}
  Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given functions.
  We say that the equation
  \begin{equation}\label{eq:filtereq}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
    -\lambda(t,x)\,w(t,x)
  \end{equation}
  is the \emph{filter equation} with \emph{driver} $X_t$, \emph{boost} $B$, and \emph{decay} $\lambda$.
\end{defn}

Filter equations afford a convenient means of computing the likelihood of a given sequence of events.
This is facilitated by the following
\begin{lemma}\label{lemma:filtereq}
  \Cref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
    \end{aligned}
  \end{equation}
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

We recognize in \cref{eq:filterlemma} the KFE of a certain process $(X_t,V_t)$.
In particular, $X_t$ is the driver with KFE \cref{eq:kfe2}.
The $V_t$ process has jumps wherever $X_t$ does, such that when $X_t$ jumps from $x$ to $x'$, $V_t$ jumps by the multiplicative factor $B(t,x,x')$.
Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
If we view $V_t$ as a weight, then \cref{lemma:filtereq} says that the $V_t$-weighted average of $X_t$ evolves according to \cref{eq:filtereq}.
This motivates the following result, which effectively allows boosts of zero.

\begin{prop}\label{prop:zeroboost}
  Suppose $X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
  Let $H_t$ be its history process.
  That is, for $\omega\in\Omega$ and $t\in\Rp$, $H_{t}(\omega):[0,t]\to\Xspace$ such that, for $t'\in{[0,t]}$,
  $H_{t}(\omega)(t')=X_{t'}(\omega)$.
  Moreover, there are random variables $K\in\Zp$ and $t_k\in\halfclosed{0,t}$, $k=1,\dots,K$ such that $\leftlim{X}_t=X_t$ whenever $t\ne{t_k}$ for all $k$.
  Suppose $F$ is a real-valued function of $H_t$ such that
  \begin{equation*}
    F(H_t)=\prod_{k}{Q\Big(t_k,\leftlim{H}(t_k),H(t_k)\Big)\,B\Big(t_k,\leftlim{H}(t_{k}),H(t_k)\Big)},
  \end{equation*}
  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
  Suppose $w$ satisfies the filter equation
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)
      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}}.
    \end{aligned}
  \end{equation*}
  Then $\Expect{F(H_t)}=\int{w(t,x)\dd{x}}$.
\end{prop}
\begin{proof}
  Apply \cref{lemma:filtereq} with the driver generated by the rate functions $\beta(t,x,x')\,Q(t,x,x')$.
\end{proof}

@@

\begin{prop}
  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function.
  Let $E$ be the set of its jump times.
  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{E}}{\delta(t,e)\delta(x',X_e)}$.
  With this driver, the filter equation (\cref{eq:filtereq}) becomes
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}=
      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\notin{E}},\\
      w(e,x)=\delta(x,X_e)\,\int{\wt(e,x')\,B(e,x',X_e)\,\dd{x'}}\quad\text{for}\quad{e\in{E}}.
    \end{gathered}
  \end{equation*}
\end{prop}

>>>>

\section{Mathematical definitions}

@@

We use bold-face to mark random elements and the corresponding regular-face symbol to denote specific values of such elements.
Given any \cadlag\ process $f_t$, the set of jump-times is denoted $\event(f_t)=\{e\in[0,t]\;\vert\;\leftlim{f}_e\ne{f_e}\}$.

@@

\subsection{The pruned and obscured genealogies}

@@

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $j\in\Zp$ and $t\in\Rp$, $Y^{P}_j(t)=\left(\ct^{P}_j(t),\anc^{P}_j(t),\deme^{P}_j(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.

@@

\paragraph{Lineage count and saturation}

@@

\paragraph{Obscured genealogy}

@@

\begin{thm}\label{thm:pruned_lik}
  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
  Then
  \begin{equation*}
    \begin{aligned}
      \Prob{P^*\vert\Hist_t}=&\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
      &\qquad\times\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}.
    \end{aligned}
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*\vert\Hist_t}=0$.
  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*\vert\Hist_t}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*$.
  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
  \end{equation*}
  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

%% \begin{equation*}
%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
%% \end{equation*}

@@

\begin{thm}\label{thm:obsc_lik}
  Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$.
  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ if and only if $Q_u(\eta,\eta')=1$.
  Let $\rdm{y}_t$ be the Markov jump process on $\Yspace$ generated by the kernel $\pi$, with $\event(\rdm{y})=\event(\Hist_t)$.
  Then
  \begin{equation*}
    \Prob{V^*\vert\Hist_t}=\Expect{\left.\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\times\Indicator{\obs(P(\rdm{y}))=V^*}\;\right\vert\;\Hist_t},
  \end{equation*}
  the expectation being taken over $\rdm{y}$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \Prob{V^*\vert\Hist_t}=\int{\Indicator{\obs(P)=V^*}\,\Prob{P\vert\Hist_t}\,\dd{P}},
  \end{equation}
  the integral being taken over all pruned genealogies.
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $V^*$ as sample paths from a Markov process on $\Yspace$ and
  treat the integral in \cref{eq:IS1} as an expectation over these paths.
  Conditional on $\Hist_t$, the probability kernel $\pi$ generates a Markov chain, $\rdm{\hat{y}}_k$, on $\Yspace$ such that
  \begin{equation*}
    \Prob{\rdm{\hat{y}}_k=\hat{y}_k\vert\rdm{\hat{y}}_{k-1}=\hat{y}_{k-1}}=\pi_{U_{t_k}}(t_k,\leftlim{X}_{t_k},X_{t_k},\hat{y}_{k-1},\hat{y}_k).
  \end{equation*}
  Now let $\event(\Hist_t)=\{t_1,t_2,\dots,t_K\}$.
  $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
  The conditions on $\pi$ guarantee that, for every pruned genealogy $P$, $Y^P$ is a sample path of this process.
  In particular, they guarantee that $\pi_u(t,\xi,\xi',\eta,\eta')\;\propto\;{Q_u(\eta,\eta')}$.
  Clearly, the construction of $\rdm{y}_t$ guarantees that $\event(P(\rdm{y}))\subseteq\event(\Hist_t)$.

  We therefore have that
  \begin{equation*}
    \Prob{V^*\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}\cdot\Indicator{\obs(P(\rdm{y}))=V^*}},
  \end{equation*}
  the expectation being taken with respect to the random process $\rdm{y}_t$.
  Here, by definition,
  \begin{equation*}
    \pi(\rdm{y}\vert\Hist_t)=\prod_{e\in\event(\Hist_t)}{\pi_{U_{e}}(e,\leftlim{X}_{e},X_{e},\leftlim{\rdm{y}}_{e},\rdm{y}_{e})}.
  \end{equation*}
  The result then follows from \cref{thm:pruned_lik}.
\end{proof}

%% \AAK{Two approaches possible:}
%% \begin{enumerate}
%% \item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
%%   In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
%% \item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
%%   That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
%%   Therefore, any importance sampling scheme for $\Prune_t$ will do.
%%   In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
%%   We show that this is equivalent to the specific filter equation.

%%   It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
%%   Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
%% \end{enumerate}

@@

\paragraph{Integrating out the history: filter equations}

@@

Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
\begin{equation}\label{eq:kfe2}
  \frac{\partial{w}}{\partial{t}}
  =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
  -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}},
\end{equation}
Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given functions.
We say that the equation
\begin{equation}\label{eq:filtereq}
  \frac{\partial{w}}{\partial{t}}
  =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
  -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
  -\lambda(t,x)\,w(t,x)
\end{equation}
is the \emph{filter equation} with \emph{driver} $X_t$, \emph{boost} $B$, and \emph{decay} $\lambda$.

Filter equations afford a convenient means of computing the likelihood of a given sequence of events.
This is facilitated by the following fact.

\begin{lemma}\label{lemma:filtereq}
  \Cref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
    \end{aligned}
  \end{equation}
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(X_t,V_t)$.
In particular, $X_t$ is the driver with KFE \cref{eq:kfe2}.
The $V_t$ process has jumps wherever $X_t$ does, such that when $X$ jumps at time $t$ from $x$ to $x'$, $V$ jumps by the multiplicative factor $B(t,x,x')>0$.
Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $V_t$-weighted average of $X_t$ evolves in time:
this average is simply $\int{w(t,x)\,\dd{x}}$.

This motivates the following result, which shows how filter equations allow one to integrate over random histories.

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
  %% Suppose that $B(t,x,x')>0$, $Q(t,x,x')\in\{0,1\}$, and $\lambda(t,x)\in\R$ are given measurable functions.
  %% Consider the equation
  %% \begin{equation*}
  %%   \frac{\partial{w}}{\partial{t}}
  %%   =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,Q(t,x',x)\,\dd{x'}}
  %%   -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
  %%   -\lambda(t,x)\,w(t,x)
  %% \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{R}_t$ is a real-valued random process such that
  \begin{equation*}
    \Expect{\rdm{R}_t\;\vert\;\Hist_t}=\prod_{e\in\event(\Hist_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
  \end{equation*}
  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
  Suppose $w$ satisfies the filter equation
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)
      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}}.
    \end{aligned}
  \end{equation*}
  Then $\Expect{\rdm{R}_t}=\int{w(t,x)\dd{x}}$.
\end{prop}
\begin{proof}
  Apply \cref{lemma:filtereq} with the driver generated by the rate functions $\beta(t,x,x')\,Q(t,x,x')$.
\end{proof}

@@

\begin{prop}
  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
  With this driver, the filter equation (\cref{eq:filtereq}) becomes
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}=
      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
      w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
    \end{gathered}
  \end{equation*}
\end{prop}
